Score-based tests of measurement invariance: Use in practice

### 2013-09-24 frontiers/frontiers.Rnw
Added comments and revisions to frontiers simulations.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@264 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

Finally, if $V$ is only nominal/categorical, the cumulative sums of scores can be used to obtain the a Lagrange multiplier statistic.
This test statistic can be formally written as
\begin{equation}
    \label{eq:lmuo}
     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
\end{equation}
and it is asymptotically equivalent to the usual, likelihood ratio test statistic.  It is advantageous over the likelihood ratio test because it requires estimation of only one model (the restricted model).  We make use of this advantage in the simulations, described later.

@@

In Simulation 1, we examined the extent to which the proposed tests
are sensitive to the specific model parameter that violates measurement
invariance.  If, say, a factor loading violates measurement
invariance, it is plausible that this violation impacts other
parameters such as factor covariances or other loadings 
associated with the factor in question.  Thus, the goal of the
simulation was to examine the extent to which the proposed tests
attribute the measurement invariance violation to the parameters that
are truly in violation.

@@

To examine these specificity issues, we generated data from a
two-factor model with three indicators each (the same model used in
Merkle and Zeileis, \citeyearNP{MerZei13}). The measurement invariance
violation occurred in one of three places: the factor loadings
 $\lambda_{11}--\lambda_{62}$, the unique variance associated
 $\epsilon_{11}--\epsilon_{66}$, or the factor covariance $\sigma_{12}$.  
We then tested for measurement invariance in seven subsets of parameters: the three
individual parameters noted above, all six factor loadings $\lambda_{11}--\lambda_{62}$, 
all six unique variances $\epsilon_{11}--\epsilon_{66}$. The path diagram is shown in  
Figure~\ref{fig:path}.
\seeme{How to refer this figure?}

@@

\begin{tikzpicture}
[latent/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=20mm, minimum height=10mm},
 manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
                inner sep=0pt,minimum width=20mm, minimum height=10mm},
 error/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=10mm},
 post/.style={->,shorten >=1pt,>=stealth',semithick},
 sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',auto,distance=20mm,semithick},
 errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',auto,distance=10mm,semithick}]

@@

\end{tikzpicture}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Power and Type I error were examined across three sample sizes
($n=120,480,960$), three numbers of categories ($m=4,8,12$) and
17 magnitudes of invariance violations. 
The measurement invariance violations began at level $1+m/2$ of $V$
and were consistent thereafter. These
violations were based on the auxiliary variable $V$: individuals below
the $1+m/2$ percentile of $V$ deviated from individuals above the $1+m/2$ 
percentile  by $d$ times the parameters' asymptotic standard errors
(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
For each combination of sample size ($n$) $\times$ violation
magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
generated and tested. Four statistics were examined above. In all conditions, we
maintained equal sample sizes at each level of the ordinal variable.

\subsection{Results}
Full simulation results are presented in Figure~\ref{fig:sim11res} to 
Figure~\ref{fig:sim13res}; all three test statistics exhibited 
similar results, with unordered  $\max\mathit{LM}$ demonstrated lower powers across all situations. ordered $\max\mathit{LM}$ and double-max exhibited equivalent power curves since there is only one parameter violating measurement invariance. Thus, different ways of aggregating $\bm B(\hat{\theta})$ would be the same.
Figure~\ref{fig:sim11res} displays power curves as a function of 
violation magnitude for the first
factor loading $\lambda_{11}$, with panels reflecting the parameters 
being tested and lines reflecting $n$.
Figures~\ref{fig:sim12res} and Figure~\ref{fig:sim13res} display 
similar power curves when the factor covariance 
$\phi_{12}$ and error variance $\epsilon_{11}$ violate measurement
invariance, respectively.

@@

In summary, we found that the proposed tests can attribute
measurement invariance violations to the correct parameter.  This
provides evidence that, in practice, one can have confidence in the
tests' abilities to locate the measurement invariance violation.  Of
course, this statement is qualified by the fact that, in this
simulation, the model was correctly specified.  In the following
simulations, we examine the tests' performance when the model is
misspecified.

@@

\seeme{I think it might be nicer to keep these plots all have three rows?(Not sure)}
\section{Simulation 2}
In Simulation 2, we examine the extent to which the results of
Simulation 1 are robust to model misspecification. Specifically, we
generate data from the factor analysis model used in the previous
section, but with adding an extra loading from the second factor to the first
indicator (of the first factor).  The estimated model lacks this extra
loading, however. The goal of this simulation is to examine the proposed statistics'
power and specificity under model misspecification.

\subsection{Method}
Measurement invariance violations could occur in
each of the three parameters from Simulation 1 (factor loading,factor covariance,unique
variance), and they could also occur in the extra, unmodeled loading. 
Sample size and magnitude of measurement invariance violation were manipulated 
in the same way as they were in Simulation 1.The tested parameters are the same as Simulation
1,too.

\subsection{Results}
Results are presented by the parameter that violated measurement invariance.
Figure~\ref{fig:sim21res} displays results when the unmodeled loading violates 
measurement invariance. One can generally observe that 
the first loading and error variance parameter showed high measurement invariance 
violation simultaneously, while factor covariance did not demonstrate violation.

When the loading $\lambda_{11}$ violates
measurement invariance, even though lacking one loading associated with the first indicator, 
statistic showed enough power to detect the first loading violation specifically. 
Similar pattern of specificity and power were observed when the violation parameter 
was $\sigma_{12}$ and $\epsilon_{11}$. Figures are shown in Appendix for simplicity.

In summary, the proposed test statistics appear robust to unmodeled loading parameters,so long
as the umodeled parameter does not violate measurment invariance. If the unmodeled parameter 
dose violate measurement invariance, however, then the tests assign this violation to other 
parameters that do not violate measurement invariance. The impacted parameters include the error
variance and other loadings associated with the manifest variables that had the unmodeled 
loading. Thus, prior to testing measurement invariance, it is important to study the extent to
which the hypothesized model includes all parameters of importance.

>>>>

% Seems necessary for texlive 2012?
\usetikzlibrary{arrows}

@@

Finally, if $V$ is only nominal/categorical, the cumulative sums of scores can be used to obtain the a Lagrange multiplier statistic.
This test statistic can be formally written as
\begin{equation}
    \label{eq:lmuo}
     \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
\end{equation}
and it is asymptotically equivalent to the usual, likelihood ratio test statistic.  It is advantageous over the likelihood ratio test because it requires estimation of only one model (the restricted model).  We make use of this advantage in the simulations, described later.

@@

In Simulation 1, we examined the extent to which the proposed tests
are sensitive to the specific model parameter that violates measurement
invariance.  If, say, a factor loading violates measurement
invariance, it is plausible that this violation impacts other
parameter estimates, including factor covariances or other loadings 
associated with the factor in question.  Thus, the goal of the
Simulation 1 is to examine the extent to which the proposed tests
attribute the measurement invariance violation to the parameters that
are truly in violation.

@@

To examine these specificity issues, we generated data from a
two-factor model with three indicators each (see Figure~\ref{fig:path}).
\readme{Please make sure the text below is correct.  The previous text stated that the violation occurs in all loadings and all unique variances.}
The measurement invariance
violation occurred in one of three places: the factor loading associated with Scale 1 
 ($\lambda_{11}$), the unique variance associated with Scale 1
 ($\epsilon_{11}$), or the factor covariance $\sigma_{12}$.  
We then tested for measurement invariance in seven subsets of parameters: the three
individual parameters noted above, all six factor loadings, % $\lambda_{11}--\lambda_{62}$, and 
and all six unique variances. % $\epsilon_{11}--\epsilon_{66}$.

@@

\begin{figure}
    \caption{General model used for the simulations.}
    \label{fig:path}
\begin{tikzpicture}
[latent/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=20mm, minimum height=10mm},
 manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
                inner sep=0pt,minimum width=20mm, minimum height=10mm},
 error/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=10mm},
 post/.style={->,shorten >=1pt,>=stealth',semithick},
 sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',auto,distance=20mm,semithick},
 errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',auto,distance=10mm,semithick}]

@@

\end{tikzpicture}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Power and Type I error were examined across three sample sizes
($n=120,480,960$), three numbers of categories ($m=4,8,12$) and
17 magnitudes of invariance violations. 
The measurement invariance violations began at level $1+m/2$ of $V$
and were consistent thereafter. These
violations were based on the auxiliary variable $V$: individuals below
the $1+m/2$ percentile of $V$ deviated from individuals above the $1+m/2$ 
percentile  by $d$ times the parameters' asymptotic standard errors
(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
For each combination of sample size ($n$) $\times$ violation
magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
generated and tested. Four statistics were examined above. \readme{What four statistics?  Equation numbers would be helpful.} 
In all conditions, we
maintained equal sample sizes at each level of the ordinal variable.

\subsection{Results}
\readme{When referring to statistics, it would be good to use the notation from Equations 6--8.  For example, instead of ``unordered $\max\mathit{LM}$'', you would have $\max\mathit{LM}_{uo}$.}
Full simulation results are presented in Figure~\ref{fig:sim11res} to 
Figure~\ref{fig:sim13res}; 
Figure~\ref{fig:sim11res} displays power curves as a function of 
violation magnitude for the first
factor loading $\lambda_{11}$, with panels reflecting the parameters 
being tested and lines reflecting $n$.
Figures~\ref{fig:sim12res} and Figure~\ref{fig:sim13res} display 
similar power curves when the factor covariance 
$\phi_{12}$ and error variance $\epsilon_{11}$ violate measurement
invariance, respectively. 
All three test statistics exhibited 
similar results, with unordered  $\max\mathit{LM}$ demonstrated lower powers across all situations. ordered $\max\mathit{LM}$ and double-max exhibited equivalent power curves because, when only one parameter is tested, the statistics are equivalent.

@@

\readme{We need to compare the specific tests: the unordered test, which mimics the LRT, has lower power.}

In summary, we found that the proposed tests can attribute
measurement invariance violations to the correct parameter.  This
provides evidence that, in practice, one can have confidence in the
tests' abilities to locate the measurement invariance violation.  Of
course, this statement is qualified by the fact that, in this
simulation, the model was correctly specified.  In the following
simulations, we examine the tests' performance when the model is
misspecified. \readme{The plots have too many points; maybe we should only plot  every other diff value.  In the xyplot command, you could add the argument \verb+subset = diff %% .5 == 0+.}

@@

\seeme{I think it might be nicer to keep these plots all have three rows?(Not sure)}

\section{Simulation 2}
In Simulation 2, we examine the extent to which the results of
Simulation 1 are robust to model misspecification. Specifically, we
generate data from the factor analysis model used in the previous
section, except that the model contains an extra loading from the second factor to Scale 1.
The estimated model matches that displayed in Figure~\ref{fig:path}, however, resulting in model misspecification.  The goal of this simulation is to examine the proposed statistics'
power and specificity under this misspecification.

\subsection{Method}
\readme{Make sure there is space after each period and comma, the beginning of each sentence is capitalized, etc.}
Measurement invariance violations could occur in
each of the three parameters from Simulation 1 (factor loading, factor covariance, and unique variance), and they could also occur in the extra, unmodeled loading.  In each condition, a single parameter exhibited the violation.
Sample size and magnitude of measurement invariance violation were manipulated 
in the same way as they were in Simulation 1.The tested parameters were also the same as Simulation
1.

\subsection{Results}
Results are presented by the parameter that violated measurement invariance.
Of primary interest is Figure~\ref{fig:sim21res}, which displays results when the unmodeled loading violates 
measurement invariance. One can generally observe that 
tests of the first loading and error variance parameter exhibited high power, reflecting a high Type I error rate.  Tests associated with the factor covariance did not demonstrate this error, however.

\readme{Need some comparison between the unordered and ordered tests.}

When the modeled factor loading, $\lambda_{11}$, violated
measurement invariance, the statistics were generally able to pick up the violation despite the misspecification.
A similar pattern of specificity and power were observed when the unique variance and factor covariance paraameters violated measurement invariance; the figures associated with these results are shown in the Appendix.

\readme{We should say something about how the unordered test looks better when the unmodeled parameter violates MI.}
In summary, the proposed test statistics appear robust to unmodeled loading parameters,so long
as the umodeled parameter does not violate measurment invariance. If the unmodeled parameter 
dose violate measurement invariance, however, then the tests assign this violation to other 
parameters that do not violate measurement invariance. The impacted parameters include the error
variance and other loadings associated with the manifest variables that had the unmodeled 
loading. Thus, prior to testing measurement invariance, it is important to study the extent to
which the hypothesized model includes all parameters of importance.
